Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-17 E-Hypergraphs-LICS24.tex
Minor fixes and S6 update

\begin{figure}
\[
	\scalebox{0.65}{
	\tikzfig{categorical-semantics/egraph-strings}
	}
\]
\caption{String diagrams for semilattice enriched symmetric monoidal categories.}
\end{figure}\label{fig:egraph-strings}
CHRIS: DONT USE PHI/PSI FOR DIAGRAMS, USE F,G

Recalling the correspondence between DAG representations of algebraic terms and  string-diagrams for morphisms of a suitable Cartesian category
makes perspicuous the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category, taken as string diagrams~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}. We extend the usual vocabulary of string diagrams for symmetric monoidal categories with the additional generator  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\]
The ability to take formal (non-empty) joins of morphisms is used to model the equivalence class structure of e-graphs.

@@

We use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n - m$ nodes,  in particular with the nodes of $m$ removed from $n$ when nodes of $m$ is a sub-e-hypergraph of $n$.
\begin{definition}[Category of e-hypergraphs with interfaces]
    The category of \textit{e-hypergraphs with interfaces} $\Ecospans$ has discrete e-hypergraphs as objects,  with homsets $\Ecospans(n,m)$ consisting of isomorphism classes of \textit{extended cospans}, defined as follows:  
    \[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
    \]
    where $\mathcal{G}$ is an e-hypergraph,  and $n, n', m, m'$ are discrete e-hypergraphs,  $f_{ext},f_{int},g_{ext},g_{int}$ are monomorphisms in $\catname{EHyp(\Sigma)}$,  and the image of $f_{ext};f_{int}$ and of $g_{ext};g_{int}$ consist exclusively of top-level vertices,  and that that vertices in the strictly internal interface (defined below) are not top-level.

@@

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessaril internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics},  we observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\qquad
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention . 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.

\begin{remark}
There is a limitation of EDPOI rewriting as defined in this section: it is unable to express rewriting
for rules whose target contains an edge with no incident vertices.  
While this limitation can be overcome,  we find the technical development (relating to the child and consistency relations) to be significantly simpler if we simply avoid consideration of such edges.  
For this reason,  in the remainder of the paper we consider signatures which do not contain $c: 0 \to 0$,  thus restricting possible e-hypergraphs to those without such problematic edges. 
%This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
%We claim that this is only a minor limitation because in practice such edges (morphisms) with no
%incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}

@@

To guarantee that the rewriting yields a monogamous directed acyclic e-hypergraph we must revise the definition of boundary complement.

@@

We are now ready to define \textit{convex extended DPOI (EDPOI) rewriting} for $\catname{EHyp({\Sigma})}$.  for e-hypergraphs with interfaces,  which we will later use to show the equivalence between $\textsf{PROP}^+(\Sigma)$ and $\WellTypedMdaEcospans$.
It is analogous to Definition~\ref{def:convex_dpo},  except we must construct internal interfaces explicitly.
%To illustrate the non-preservation of internal interfaces, consider the rewrite rule above and the following e-hypergraph with interfaces to perform the rewrite on
% \[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
% such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
%\[
%    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k
%\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$  and assume $i' \not \cong i''$ and $j' \not \cong j''$.

@@

CHRIS: TODO: COMBINE TWO EDPOI DEFINITIONS
\begin{definition}[Convex EDPOI rewriting]

\label{def:dpoi-e}
Given a span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites to} $\mathcal{H}$ \textit{with external interface} $n + k$ \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\]
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]

    Let $\mathfrak{R}$ be a set of EDPOI rewrite rules.
    Then, given 
    \[
        \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k
    \quad 
    \text{and} 
    \quad
        \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
    \] in $\catname{EHyp(\Sigma)}$ with the external interfaces $n + k$, $\mathcal{G}$ \textit{rewrites convexly into} $\mathcal{H}$ \textit{with the outermost interface} $n + k$ --- notation $( \mathcal{G} \xleftarrow{} n + k) \Rrightarrow_{\mathfrak{R}}^{+} (\mathcal{H} \xleftarrow{} n + k)$ --- if there exists a rule 
    \[
        \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
    \] 
    with the external interfaces $i + j$ in $\mathfrak{R}$ and object $\mathcal{C}$ and cospan arrows $i + j \xrightarrow{} \mathcal{C} \xleftarrow{} n + k$ 
    such that the diagram of Definition~\ref{def:dpoi-e} commutes and its marked squares are pushouts, and the following conditions hold

    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item all $v,w$ that have a pre-image in $i+j$ share the same set of parents;
        \item $m(v)  \consistency m(w)$ for all $v,w$ that have a pre-image in $i+j$.
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement in the leftmost pushout.
    \end{enumerate}
and
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
\begin{align*}
&n'' = (i'' \setminus i) + l\\
&k'' = (j'' \setminus j) + r\\
&n'' \to \mathcal{H} = [f_1,f_2] \text{ where }\\
&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H}\\ 
&f_2 : l \to \mathcal{C} \to \mathcal{H} 
\end{align*}
Since both $f_1$ and $f_2$ are monos, their copairing is also a mono. 
Similarly for $k'' \to \mathcal{H}$.
\end{definition}
The conditions (3) and (4) on a match, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]

With our modified definition of extended DPOI rewriting in hand,  we proceed to prove that our combinatorial representation of semilattice-enriched SMCs is sound and complete in an appropriate sense. 
%\begin{proposition}
%    $n \xrightarrow{} n'' \xrightarrow{} \mathcal{H} \xleftarrow{} k'' \xleftarrow{} k$ is a well-typed mda-cospan.
%\end{proposition}

@@

We wish to specify a set of EDPOI rewrite rules corresponding to those in Figure \ref{fig:string-equations}.  In order to do this,  we specify a function which maps  uninterpreted $\Sigma^+$-terms into $\WellTypedMdaEcospans$ morphisms,  and which also extends the expected interpretation of $\Sigma$-terms in $\WellTypedMdaEcospans$ (as given by the free functor from $\textbf{PROP}(\Sigma)$ induced by the interpretation of $\Sigma$ as itself,  detailed in \cite{bonchi_string_2022-2}). 
\begin{definition}[Interpretation function] 
%Let $\llbracket - \rrbracket: \textbf{PROP}(\Sigma) \to \WellTypedMdaEcospans$ be the unique SMC-functor induced by the interpretation sending $\Sigma$ to itself.  
Let $\llbracket - \rrbracket_{n,m}: \Sigma^+(n,m) \to \WellTypedMdaEcospans(n,m)$ be the function defined by induction on the type derivation of $\Sigma$-terms by 
\begin{align*}
	\llbracket \id \rrbracket_{n,m} &= \id  &	\llbracket \phi ; \psi \rrbracket_{n,m} &= \llbracket \phi \rrbracket_{n,m} ; \llbracket \psi \rrbracket_{n,m} \\
	\llbracket 1 \rrbracket_{n,m} &= 1 & \llbracket f \otimes g \rrbracket_{n,m} &= \llbracket f \rrbracket^+_{n,m} \otimes \llbracket g \rrbracket_{n,m} \\
	\llbracket \sym \rrbracket_{n,m} &= \sym  & \llbracket c \rrbracket_{n,m} &= c   \\ 
	\llbracket f+g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} + \llbracket g \rrbracket^+_{n,m}
\end{align*}
where $c \in \Sigma$ and $+$ in $\WellTypedMdaEcospans$ is defined by Figure \ref{fig:A+B}.  We will omit the subscripts in $\llbracket - \rrbracket_{n,m}$ where it is clear from context. 
\end{definition}

>>>>

\begin{figure}
\[
	\scalebox{0.65}{
	\tikzfig{categorical-semantics/egraph-strings}
	}
\]
\caption{String diagrams for semilattice enriched symmetric monoidal categories.}
\end{figure}\label{fig:egraph-strings}

Recalling the correspondence between DAG representations of algebraic terms and  string-diagrams for morphisms of a suitable Cartesian category
makes perspicuous the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category, taken as string diagrams~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}. We extend the usual vocabulary of string diagrams for symmetric monoidal categories with the additional generator  
\[
\infer{f_1 + \ldots + f_n: A \to B}{\{f_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\]
The ability to take formal (non-empty) joins of morphisms is used to model the equivalence class structure of e-graphs.

@@

We use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n - m$ nodes,  in particular with the nodes of $m$ removed from $n$ when nodes of $m$ is a sub-e-hypergraph of $n$.
\begin{definition}[Category of e-hypergraphs with interfaces]
    The category of \textit{e-hypergraphs with extended interfaces} $\Ecospans$ has discrete e-hypergraphs as objects,  with homsets $\Ecospans(n,m)$ consisting of isomorphism classes of \textit{extended cospans}, defined as follows:  
    \[
    n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
    \]
    where $\mathcal{G}$ is an e-hypergraph,  and $n, n', m, m'$ are discrete e-hypergraphs,  $f_{ext},f_{int},g_{ext},g_{int}$ are monomorphisms in $\catname{EHyp(\Sigma)}$,  and the image of $f_{ext};f_{int}$ and of $g_{ext};g_{int}$ consist exclusively of top-level vertices,  and that that vertices in the strictly internal interface (defined below) are not top-level.  We will sometimes write $\mathcal G$ to denote the extended cospan,  where it is clear from context $\mathcal G$ is equipped with extended interfaces.

@@

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics},  we observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\qquad
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention . 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.

Before introducing our definition of \textit{extended} DPOI rewriting,  we must modify the definition of boundary complement to guarantee that the rewriting yields a monogamous directed acyclic e-hypergraph.

\begin{remark}
There is a limitation of extended DPOI rewriting as defined in this section: it is unable to express rewriting
for rules whose target contains an edge with no incident vertices.  
While this limitation can be overcome,  we find the technical development to be significantly simpler if we simply avoid consideration of such edges.  
For this reason,  in the remainder of the paper we consider signatures which do not contain $c: 0 \to 0$,  thus restricting possible e-hypergraphs to those without such problematic edges. 
%This is because there is no way to impose both consistency and child relation of such an edge as it has no incident nodes in its context.
%We claim that this is only a minor limitation because in practice such edges (morphisms) with no
%incident nodes correspond to ‘scalars’, if we consider for example a category of vector spaces, and scalars are usually removed by being absorbed by vectors rather than introduced during rewrites.
\end{remark}

@@

@@

We are now ready to define \textit{convex extended DPOI (EDPOI) rewriting} for $\catname{EHyp({\Sigma})}$.  
It is analogous to Definition~\ref{def:convex_dpo},  except we must construct internal interfaces explicitly.
%To illustrate the non-preservation of internal interfaces, consider the rewrite rule above and the following e-hypergraph with interfaces to perform the rewrite on
% \[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
% such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
%\[
%    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k
%\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$  and assume $i' \not \cong i''$ and $j' \not \cong j''$.

@@

CHRIS: FIX BELOW DEFINITION. 
\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
 and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item all elements in the image of $i+j$ share the same set of parents;
        \item $m(v)  \consistency m(w)$ for all $v,w$ in the image of $i+j$.
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces are given by the following:
		\begin{align*}
		&n'' = (i'' \setminus i) + l\\
		&k'' = (j'' \setminus j) + r\\
		&n'' \to \mathcal{H} = [f_1,f_2] \text{ where }\\
		&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H}\\ 
		&f_2 : l \to \mathcal{C} \to \mathcal{H} 
		\end{align*}
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
Since both $f_1$ and $f_2$ are monos, their copairing is also a mono. 
Similarly for $k'' \to \mathcal{H}$.

    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The conditions (3) and (4) on a match, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]

With our modified definition of extended DPOI rewriting in hand,  we proceed to prove that our combinatorial representation of semilattice-enriched SMCs is sound and complete in an appropriate sense.

%\begin{proposition}
%    $n \xrightarrow{} n'' \xrightarrow{} \mathcal{H} \xleftarrow{} k'' \xleftarrow{} k$ is a well-typed mda-cospan.
%\end{proposition}

@@

We wish to specify a set of EDPOI rewrite rules corresponding to those in Figure \ref{fig:string-equations}.  In order to do this,  we specify a function which maps  uninterpreted $\Sigma^+$-terms into $\WellTypedMdaEcospans$ morphisms,  and which also extends the expected interpretation of $\Sigma$-terms in $\WellTypedMdaEcospans$ (as given by the free functor from $\textbf{PROP}(\Sigma)$ induced by the interpretation of $\Sigma$ as itself,  detailed in \cite{bonchi_string_2022-2}). 
\begin{definition}[Interpretation function] 
%Let $\llbracket - \rrbracket: \textbf{PROP}(\Sigma) \to \WellTypedMdaEcospans$ be the unique SMC-functor induced by the interpretation sending $\Sigma$ to itself.  
Let $\llbracket - \rrbracket_{n,m}: \Sigma^+(n,m) \to \WellTypedMdaEcospans(n,m)$ be the function defined by induction on the type derivation of $\Sigma$-terms by 
\begin{align*}
	\llbracket \id \rrbracket_{n,m} &= \id  &	\llbracket f ; g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} ; \llbracket g \rrbracket_{n,m} \\
	\llbracket 1 \rrbracket_{n,m} &= 1 & \llbracket f \otimes g \rrbracket_{n,m} &= \llbracket f \rrbracket^+_{n,m} \otimes \llbracket g \rrbracket_{n,m} \\
	\llbracket \sym \rrbracket_{n,m} &= \sym  & \llbracket c \rrbracket_{n,m} &= c   \\ 
	\llbracket f+g \rrbracket_{n,m} &= \llbracket f \rrbracket_{n,m} + \llbracket g \rrbracket^+_{n,m}
\end{align*}
where $c \in \Sigma$ and $+$ in $\WellTypedMdaEcospans$ is defined by Figure \ref{fig:A+B}.  We will omit the subscripts in $\llbracket - \rrbracket_{n,m}$ where it is clear from context. 
\end{definition}
